6.829 Final Project: A comparison of bit-rate selection algorithms

### 2013-05-16 paper/paper.tex
TeX cleanup

% TEMPLATE for Usenix papers, specifically to meet requirements of
%  USENIX '05
% originally a template for producing IEEE-format articles using LaTeX.
%   written by Matthew Ward, CS Department, Worcester Polytechnic Institute.
% adapted by David Beazley for his excellent SWIG paper in Proceedings,
%   Tcl 96
% turned into a smartass generic template by De Clarke, with thanks to
%   both the above pioneers
% use at your own risk.  Complaints to /dev/null.
% make it two column with no page numbering, default is 10 point

% Munged by Fred Douglis <douglis@research.att.com> 10/97 to separate
% the .sty file from the LaTeX source template, so that people can
% more easily include the .sty file into an existing document.  Also
% changed to more closely follow the style guidelines as represented
% by the Word sample file.

% Note that since 2010, USENIX does not require endnotes. If you want
% foot of page notes, don't include the endnotes package in the 
% usepackage command, below.

% This version uses the latex2e styles, not the very ancient 2.09 stuff.
\documentclass[letterpaper,twocolumn,10pt]{article}
\usepackage{epsfig,graphicx,usenix,fullpage, hyperref}

%\usepackage{endnotes}
\begin{document}

%don't want date printed

%make title bold and 14 pt font (Latex default is non-bold, 16 pt)
\title{\Large \bf 6.829 Final Project: A comparison of bit-rate selection algorithms}
%for single author (just remove % characters)
\author{
{\rm Colleen Josephson}\\
cjoseph@mit.edu
\and
{\rm Pavel Panchekha}\\
pavpan@mit.edu
% copy the following lines to add more authors
% \and
% {\rm Name}\\
%Name Institution
} % end author

\date{May 2013}

\maketitle

% Use the following at camera-ready time to suppress page numbers.
% Comment it out when you first submit the paper for review.
%\thispagestyle{empty}

\subsection*{Abstract}
In this paper we compare the performance SampleRate and Minstrel, two
popular bit-rate selection algorithms that have widespread real-world
usage on the Linux operating system. We use a novel trace-based
approach that allows us avoid in-depth kernel programming and write
simple Python scripts instead. We also introduce improvements to the
Minstrel algorithm that allow for significant gains in throughput.
\section{Introduction}

One of the key ways that wireless networks differ from wired is that
wireless networks have varying link rates. Link conditions vary with
time due to interference from other devices, changing network
geometry, and mobile clients. An optimal rate at some time $t$ may be
different from the optimal rate just 30 seconds earlier. A good
bit-rate selection algorithm has to detect and adapt to these
conditions. If the chosen bit-rate is too slow, then the throughput
will be unnecessarily low. If the rate is too high, then failures will
be very frequent and throughput will again suffer.

@@

SampleRate has three main functions: \texttt{ApplyRate()}, which
returns a the bit-rate to send a packet at,
\texttt{ProcessFeedback()}, which updates the statistics for a bitrate
after an attempted packet send, and \texttt{RemoveStaleResults()},
which removes all statistics older than 10 seconds.

@@

Our initial plan was to use the algorithms as implemented in the
MadWifi drivers for the Atheros chipset. However, MadWifi is
deprecated, and only runs on now-ancient network cards. In addition to
the unavailability of compatible hardware, MadWifi would not be an
accurate reflection of contemporary users. Both ath5k and ath9k, which
were created to replace MadWifi, did not port the old bit-rate
selection system from MadWifi and instead opted to use the Linux
kernel's rate selection framework, which only implements Minstrel.
Porting SampleRate over to the new drivers would not have been
trivial, as there are vast differences in the assumed interfaces.

@@

\begin{figure}[hbt]
  \centering
  \includegraphics[width=3.5in]{bsmt13floorplan.png}
  \caption{We collected most traces at a remote access point in the basement of Building 13 at MIT.}
\label{figure:1}
\end{figure}

@@

In Figure~\ref{figure:2} we show the performance of constant bitrates
on the \texttt{clear\_1.dat} trace, which is a 35 second trace in
clear view of the access point. The optimal rate here was 18 Mbps,
which achieved an average throughput of about 15 Mbps. The next
highest bitrate, 24 Mbps, shows a large plummet in performance, down
to a throughput of 1.17 Mbps. This steep drop is charateristic, and is
also talked about in the SampleRate thesis~\cite{samplerate}. It
demonstrates just how important bit rate selection is---the optimal
bit rate has typically much higher throughput than the next-best rate,
but if the algorithm strays too high the results can be devastating.

@@

The histograms of which rates the algorithms decided to send at
provided some useful insight into their behavior. Table~\ref{table:2}
is a histogram of a 35 second corner trace. In this situation Minstrel
outperforms SampleRate. The constant rate throughputs demonstrate that
11, 12 and 18 Mbps were the optimal rates to send at, and that no
packets were sucessful at any higher rates. SampleRate does not use
these rates except for probe packets, which happens once every ten
seconds. Minstrel, since it does not terminate transmission after four
successive failures, spends a lot of time sending probe packets at
these high rates, which are never successful. The only reason low
rates have fewer probe packets is because Minstrel explicitly places
them second in the retry chain, and because the main rate usually
succeeds. Despite the fact that SampleRate spends less time probing
high rates, it is less successful. This is because Minstrel send the
most packets at the three most successful bitrates. SampleRate,
however, spends too much time sampling low bit rates. Additionally, it
spends a lot of time sending at 5.5 Mbps. TODO: why does SampleRate
spend time at 5.5 Mbps??? We suspect that this might be because the
equation SampleRate uses to estimate average transmission time needs
to be tweaked.

@@

\begin{figure}[htb]
  \hspace{-1em}\includegraphics[width=3.4in]{constant.png}\vspace{-0em}
  \caption{Throughput of using a constant 802.11b/g bitrate in clear line-of-sight from the access point.}
\label{figure:2}
\end{figure}

\begin{figure}[htb]
  \hspace{-.75em}\includegraphics[width=3.5in]{MinVSSam1.png}\vspace{-0.75em}
  \caption{Throughput of Minstrel and SampleRate on various trace situations.}
\label{figure:3}
\end{figure}

@@

@@

However, Minproved still makes many poor choices.  While a single
successful probe packet does not cause many hundreds of milliseconds
of failed packets, as it does in Minstrel, the single successful
packet can still cause one window's worth of failed packets.  Instead,
it would be best if a successful probe packet at a rarely-successful
bit-rate was treated more carefully.  One can imagine an algorithm,
for example, which would treat a single successful probe packet as a
reason to send a dozen

\begin{table*}[htb]
    \centering
    \begin{tabular}{l|l|l|l|l}
    \textbf{Rate in Mbps}     & \textbf{SampleRate} & \textbf{Minstrel}   & \textbf{Minproved}   & \textbf{Constant Rate Throughput}\\ \hline
    1                & 631        & 39         & 29      & 0.639 Mbps    \\
    2                & 863        & 82         & 20      & 1.425 Mbps    \\
    5.5              & 4814       & 110        & 805     & 4.406 Mbps    \\
    6                & 423       & 137        & 276     & 4.603 Mbps    \\
    9                & 553         & 327        & 57      & 4.630 Mbps   \\
    11               & 3326       & 4121       & 3356    & 9.627 Mbps   \\
    12               & 4649       & 9515       & 14285   & 9.444 Mbps   \\
    18               & 1154         & 13295      & 15795   & 8.458 Mbps    \\
    24               & 16         & 6909       & 1918    & 0 Mbps   \\
    36               & 16         & 6944       & 2030    & 0 Mbps   \\
    48               & 16         & 6888       & 2009    & 0 Mbps   \\
    54               & 16         & 7308       & 1953    & 0 Mbps   \\ \hline
    \textbf{Avg. Throughput:} & \emph{5.42 Mbps}  & \emph{7.578 Mbps} & \emph{11.107 Mbps} \\
    \end{tabular}
    \caption{Histograms of the number of packets sent at each rate.
      The trace was \texttt{corner\_1.dat} and was recorded around the
      corner from the access point. The best fixed rate for this trace
      was 11 Mbps, which achieved a throughput of 9.627
      Mbps. Minproved achieves a higher throughput than the best fixed
      rate.}
\label{table:2}
\end{table*}

\begin{figure}[htb]
  \hspace{-2em}\includegraphics[width=3.5in]{mnVSspVSmp1.png}\vspace{-0em}
  \caption{Throughput of Minproved as compared to SampleRate and vanilla Minstrel.}
\label{figure:4}
\end{figure}

>>>>

% TEMPLATE for Usenix papers originally for producing IEEE-format articles
% written by Matthew Ward, CS Department, Worcester Polytechnic Institute.
% adapted by David Beazley for his paper in Proceedings, Tcl 96
% turned into a smartass generic template by De Clarke

\documentclass[twocolumn,10pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx,hyperref}
\usepackage{subcaption}

\title{6.829 Final Project: A comparison of bit-rate selection algorithms}
\author{Colleen Josephson \\ \texttt{cjoseph@mit.edu}
  \and Pavel Panchekha \\ \texttt{pavpan@mit.edu}}
\date{May 2013}

\begin{document}
\maketitle

\begin{abstract}
We compare the performance SampleRate and Minstrel, two popular
bit-rate selection algorithms that have widespread real-world usage.
We use a trace-based approach to avoid kernel programming and consider
Python implementations of both SampleRate and Minstrel.  We test both
algorithms in multiple real-world scenarios, including scenarios with
mobile clients and noisy environments, to highlight differences
between the two algorithms.  We also introduce improvements to the
Minstrel algorithm that allow for significant gains in throughput.
\end{abstract}

\section{Introduction}

One of the key ways that wireless networks differ from wired links is
that wireless networks have varying link rates.  Link conditions vary
with time due to interference from other devices, changing network
geometry, and mobile clients.  An optimal rate at one time may be
different from the optimal rate just 30 seconds earlier.  A good
bit-rate selection algorithm has to detect and adapt to these
conditions.  If the chosen bit-rate is too slow, then the throughput
will be unnecessarily low; if the rate is too high, failures will
be very frequent and throughput will again suffer.

@@

SampleRate has three main functions: \texttt{ApplyRate()}, which
returns a the bit-rate to send a packet at,
\texttt{ProcessFeedback()}, which updates the statistics for a bit-rate
after an attempted packet send, and \texttt{RemoveStaleResults()},
which removes all statistics older than 10 seconds.

@@

Our initial plan was to use the algorithms as implemented in the
MadWifi drivers for the Atheros chip.  However, MadWifi is
deprecated, and only runs on now-ancient network cards.  In addition to
the unavailability of compatible hardware, MadWifi would not be an
accurate reflection of contemporary users.  Both ath5k and ath9k, which
were created to replace MadWifi, did not port the old bit-rate
selection system from MadWifi and instead opted to use the Linux
kernel's rate selection framework, which only implements Minstrel.
Porting SampleRate over to the new drivers would not have been
trivial, as there are vast differences in the assumed interfaces.

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\linewidth]{bsmt13floorplan.png}
  \caption{We collected most traces at a remote access point in the basement of Building 13 at MIT.}
\label{figure:1}
\end{figure}

@@

\begin{figure*}[htbp]
  \centering
  \begin{subfigure}{.45\linewidth}
    \includegraphics[width=\linewidth]{constant.png}
    \caption{Throughput of using a constant 802.11b/g bit-rate in
      clear line-of-sight from the access point.}
    \label{figure:2}
  \end{subfigure}\hfill
  \begin{subfigure}{.45\linewidth}
    \includegraphics[width=\linewidth]{MinVSSam1.png}
    \caption{Throughput of Minstrel and SampleRate in various
      scenarios.}
    \label{figure:3}
  \end{subfigure}
  \caption{Comparing Minstrel and Samplerate}
\end{figure*}

In Figure~\ref{figure:2} we show the performance of constant bit-rates
on the \texttt{clear\_1.dat} trace, which is a 35 second trace in
clear view of the access point.  The optimal rate here was 18 Mbps,
which achieved an average throughput of about 15 Mbps.  The next
highest bit-rate, 24 Mbps, shows a large plummet in performance, down
to a throughput of 1.17 Mbps.  This steep drop is charateristic, and is
also talked about in the SampleRate thesis~\cite{samplerate}.  It
demonstrates just how important bit rate selection is---the optimal
bit rate has typically much higher throughput than the next-best rate,
but if the algorithm strays too high the results can be devastating.

@@

The histograms of which rates the algorithms decided to send at
provided some useful insight into their behavior.  Table~\ref{table:2}
is a histogram of a 35 second corner trace.  In this situation
Minstrel outperforms SampleRate.  The constant rate throughputs
demonstrate that 11, 12 and 18 Mbps were the optimal rates to send at,
and that no packets were sucessful at any higher rates.  SampleRate
does not use these rates except for probe packets, which happens once
every ten seconds.  Minstrel, since it does not terminate transmission
after four successive failures, spends a lot of time sending probe
packets at these high rates, which are never successful.  The only
reason low rates have fewer probe packets is because Minstrel
explicitly places them second in the retry chain, and because the main
rate usually succeeds.  Despite the fact that SampleRate spends less
time probing high rates, it is less successful.  This is because
Minstrel send the most packets at the three most successful bit-rates.
SampleRate, however, spends too much time sampling low bit rates.
Additionally, it spends a lot of time sending at 5.5 Mbps.  We suspect
that this might be because the equation SampleRate uses to estimate
average transmission time needs to be tweaked.
%   TODO: why does SampleRate spend time at 5.5 Mbps???

@@

@@

\begin{table*}[htb]
    \centering
    \begin{tabular}{l|l|l|l|l}

    \textbf{Rate in Mbps} & \textbf{SampleRate} &
    \textbf{Minstrel} & \textbf{Minproved} &
    \textbf{Constant Rate Throughput}\\ \hline

    1   & 631  & 39    & 29    & 0.639 Mbps \\
    2   & 863  & 82    & 20    & 1.425 Mbps \\
    5.5 & 4814 & 110   & 805   & 4.406 Mbps \\
    6   & 423  & 137   & 276   & 4.603 Mbps \\
    9   & 553  & 327   & 57    & 4.630 Mbps \\
    11  & 3326 & 4121  & 3356  & 9.627 Mbps \\
    12  & 4649 & 9515  & 14285 & 9.444 Mbps \\
    18  & 1154 & 13295 & 15795 & 8.458 Mbps \\
    24  & 16   & 6909  & 1918  & 0 Mbps \\
    36  & 16   & 6944  & 2030  & 0 Mbps \\
    48  & 16   & 6888  & 2009  & 0 Mbps \\
    54  & 16   & 7308  & 1953  & 0 Mbps \\ \hline

    \textbf{Avg. Throughput:} & \emph{5.42 Mbps}  & \emph{7.578 Mbps} & \emph{11.107 Mbps} \\
    \end{tabular}
    \caption{Histograms of the number of packets sent at each rate.
      The trace was \texttt{corner\_1.dat} and was recorded around the
      corner from the access point.  The best fixed rate for this
      trace was 11 Mbps, which achieved a throughput of 9.627 Mbps.
      Minproved achieves a higher throughput than the best fixed
      rate.}

    \label{table:2}
\end{table*}

\begin{figure}[htb]
  \includegraphics[width=\linewidth]{mnVSspVSmp1.png}
  \caption{Throughput of Minproved as compared to SampleRate and
    vanilla Minstrel.}
\label{figure:4}
\end{figure}

@@

However, Minproved still makes many poor choices.  While a single
successful probe packet does not cause many hundreds of milliseconds
of failed packets, as it does in Minstrel, the single successful
packet can still cause one window's worth of failed packets.  Instead,
it would be best if a successful probe packet at a rarely-successful
bit-rate was treated more carefully.  One can imagine an algorithm,
for example, which would treat a single successful probe packet as a
reason to send a dozen more probe packets at that bit-rate, but does
not yet commit to sending hundreds of packets at that bit-rate.

However, these are only vague ideas, and we have no concrete
implementations of such an algorithm.
